Estimation of Stochastic Wilson and Cowan Neural Field Model using Expectation Maximisation Algorithm

### 2011-05-16 ltx/data_paper/DataPaper.tex
minor edits
minor edit to results

\boldsymbol\Xi_{1}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right]\label{eq:defofXi1} \\
\boldsymbol\Xi_{2}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t}^\top\right]\\

@@

	\boldsymbol\Xi_{0}&=\sum_{t=0}^{T-1}\left(\mathbf P_{t+1}^b+\mathbf{\hat x}_{t+1}^b\mathbf{\hat x}_{t+1}^{b\top}\right)\label{eq:defofXi0}, \\
\boldsymbol\Xi_{1}&=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}^b+\mathbf{\hat x}_t^b\mathbf{\hat x}_{t+1}^{b\top}\right) \label{eq:defofXi1}, \\
\boldsymbol\Xi_{2}&=\sum_{t=0}^{T-1}\left(\mathbf P_t^b+\mathbf{\hat x}_t^b\mathbf{\hat x}_t^{b\top}\right).

@@

\boldsymbol\Xi_{3}&\approx\sum_{t=0}^{T-1}\left[ \mathbf{\hat x}_{t+1}^{b\top}\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf{\hat x}_t^b)+\int_\Omega\boldsymbol \phi^\top(\mathbf r') \mathbf M_{t+1}^b \tilde{\boldsymbol\Sigma}_e^{-1} \boldsymbol{\Psi}(\mathbf{r}') f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t^b) d\mathbf{r}'\right] \\	
\boldsymbol\Xi_{4}&\approx\sum_{t=0}^{T-1}\left[q^\top(\mathbf{\hat x}_t^b)\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf{\hat x}_t^b)+\Lambda_t\right] \\
 \boldsymbol\Xi_{5}&\approx\sum_{t=0}^{T-1}\left[ \mathbf{\hat x}_{t}^{b\top}\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf{\hat x}_t^b)+\int_\Omega\boldsymbol \phi^\top(\mathbf r') \mathbf P_t^b \tilde{\boldsymbol\Sigma}_e^{-1}  \boldsymbol{\Psi}(\mathbf{r}') f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t^b) d\mathbf{r}'\right] 

@@

@@

\begin{equation}\label{eq:theta}

@@

@@

@@

Variable &Order&Order of the approximate\\ \hline\\
$\Xi_0, \Xi_1, \Xi_2$&$O(Tn_x^2)$&-\\
$\Xi_3, \Xi_5$&$O(Tn_x^2n_{\theta})$&$O(Tn_x^3)+O(Tn_x^2n_r)+O(Tn_xn_{\theta}n_r)$ \\
$\Xi_4$&$O(Tn_x^2n_{\theta})+O(Tn_xn_{\theta}^2)$&$O(Tn_x^4n_{\theta})+O(Tn_x^3n_{\theta}^2)+O(Tn_xn_{\theta}n_r)$\\
$\hat{\boldsymbol\theta} $&$O(n_{\theta}^3)$&-\\
$\hat{\xi}$&$O(n_x^3+n_{\theta}^3)$&-\\
 $\hat{\sigma}_{\epsilon}^2$&$O(Tn_y^2n_x^2)$&-\\
$\hat{\sigma}_{d}^2$&$O(Tn_x^3)+O(Tn_{\theta}^3)$&-\\

@@

The correlation analysis described in Section \ref{subsec:CorrAnal} were applied to generated data in order to approximate the unknown supports of both connectivity kernel and the disturbance covariance function. In the algorithm, the unknown variables $\sigma_{\epsilon}^2$ and $\xi$ in equations () and () were set to zero. By doing so, the exact shape of the kernel and disturbance covariance function cannot be inferred, this step might be considered as an initialisation for the EM algorithm, however, an approximate of their support can be still extracted. In fact, the estimated support for the disturbance covariance function was very accurate. The estimates of the connectivity kernel and disturbance support were normalised and thresholded for values less than 0.01. The result of the correlation analysis for the connectivity kernel is shown in Figure.\ref{fig:KernelWidthEstimation}. The red circle shows the estimated support which encompasses the non-zero values (black area) after thresholding. The estimated support of the disturbance covariance function convolved twice with the observation kernel is shown in Figure.\ref{fig:DisturbanceWidthEstimation}. By fitting a Guassian basis function to the result, and under the assumption that the width of the sensor,$\sigma_m^2$ were known, the width of the disturbance covariance function, $\sigma_d^2$ was calculated analytically using the expression for Gaussian basis functions convolution.

The estimated supports were then applied in the EM algorithm described in Section \ref{sec:EstimationAlgorithm} to estimate the smoothed states $\hat{\mathbf x}_t^b$, the connectivity kernel weights $\boldsymbol\theta$, the synaptic dynamic $\xi$, and the hyperparameters $\sigma_{\epsilon}^2$ and $\sigma^2_d$. A  square grid of $n_{\theta}=25$ (5 by 5) basis functions with $2.25~mm$ spacing were used to cover the estimated support where $\sigma_{\psi}^2=3.5$. This is shown in Figure.\ref{fig:KernelWidthEstimation} by blue circles, each of which representing full width at half maximum of the connectivity kernel basis functions. The true and estimated connectivity kernel for 100 Monte Carlo simulations are shown in Figure~\ref{fig:KernelEstimates}. Cross-sections of the connectivity kernel estimates for each run are shown in Figure~\ref{fig:KernelEstimateCrossSection}. The reconstructed kernel is in good accordance with the actual kernel, where the actual kernel is inside the confidence interval. The large standard deviation is due to the high number of parameters need to be estimated. Figure~\ref{fig:Histogram} depicts the histogram of parameter estimates from 100 simulations where the true parameter value is shown by the solid line and the mean parameter estimate is shown by the dashed line. These histograms indicate small bias in parameter estimates with a very small standard deviation. Examples of neural field reconstruction are given in Figure.~\ref{fig:FieldEstimates}.  The figure shows snapshots at three time instants of the true neural field and the estimated field. In each subplot neural activity is plotted against spatial index in 1D to provide a comparision between true and the estimated underlying fields. Plot of lower bound of the log likelihood versus iterations of the EM is shown in Figure.~\ref{fig:LowerBound}. In general, for the non-linear dynamical systems the actual lower bound can not be claculated analytically, therefore the approximate of the lower bound obtained by URTSS is shown. The change in lower bound drops below 0.009 percent after 15 iterations, confirming
that the algorithm converged to steady parameter values.

\begin{itemize}
	\item Will show parameter distributions
	\item Will show lower bound
      \end{itemize}

\section{Discussion}

>>>>

\boldsymbol\Xi_{1}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right]\label{eq:Xi1} \\
\boldsymbol\Xi_{2}&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t}^\top\right]\label{eq:Xi2}\\

@@

	\boldsymbol\Xi_{0}&=\sum_{t=0}^{T-1}\left(\mathbf P_{t+1}^b+\mathbf{\hat x}_{t+1}^b\mathbf{\hat x}_{t+1}^{b\top}\right)\label{eq:Xi0Gibson}, \\
\boldsymbol\Xi_{1}&=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}^b+\mathbf{\hat x}_t^b\mathbf{\hat x}_{t+1}^{b\top}\right) \label{eq:Xi1Gibson}, \\
\boldsymbol\Xi_{2}&=\sum_{t=0}^{T-1}\left(\mathbf P_t^b+\mathbf{\hat x}_t^b\mathbf{\hat x}_t^{b\top}\right).\label{eq:Xi2Gibson}

@@

\boldsymbol\Xi_{3}&\approx\sum_{t=0}^{T-1}\left[ \mathbf{\hat x}_{t+1}^{b\top}\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf{\hat x}_t^b)+\int_\Omega\boldsymbol \phi^\top(\mathbf r') \mathbf M_{t+1}^b \tilde{\boldsymbol\Sigma}_e^{-1} \boldsymbol{\Psi}(\mathbf{r}') f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t^b) d\mathbf{r}'\right] \label{eq:Xi3Ap}\\	
\boldsymbol\Xi_{4}&\approx\sum_{t=0}^{T-1}\left[q^\top(\mathbf{\hat x}_t^b)\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf{\hat x}_t^b)+\Lambda_t\right]\label{eq:Xi4Ap} \\
 \boldsymbol\Xi_{5}&\approx\sum_{t=0}^{T-1}\left[ \mathbf{\hat x}_{t}^{b\top}\tilde{\boldsymbol\Sigma}_e^{-1}q(\mathbf{\hat x}_t^b)+\int_\Omega\boldsymbol \phi^\top(\mathbf r') \mathbf P_t^b \tilde{\boldsymbol\Sigma}_e^{-1}  \boldsymbol{\Psi}(\mathbf{r}') f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t^b) d\mathbf{r}'\right] \label{eq:Xi5Ap}

@@

\begin{equation}\label{eq:xihat}

@@

\begin{equation}\label{eq:thetahat}

@@

\begin{equation}\label{eq:sigmaepsilonhat}

@@

\begin{align}\label{eq:sigmadhat}

@@

Variable &Equation&Order\\ \hline\\
$\Xi_0, \Xi_1, \Xi_2$&\eqref{eq:Xi0Gibson}, \eqref{eq:Xi1Gibson}, \eqref{eq:Xi2Gibson}&$O(Tn_x^2)$\\
$\Xi_3, \Xi_5$&\eqref{eq:Xi3Ap},\eqref{eq:Xi5Ap} &$O(Tn_x^3)+O(Tn_x^2n_r)+O(Tn_xn_{\theta}n_r)+O(Tn_{\theta}n_r^2)$ \\
$\Xi_4$&\eqref{eq:Xi4Ap}&$O(Tn_x^4n_{\theta})+O(Tn_x^3n_{\theta}^2)+O(Tn_x^3n_{\theta}n_r^2)$\\
$\hat{\xi}$&\eqref{eq:xihat}&$O(n_x^3+n_{\theta}^3)$\\
$\hat{\boldsymbol\theta} $&\eqref{eq:thetahat}&$O(n_{\theta}^3)$\\
$\hat{\sigma}_{\epsilon}^2$&\eqref{eq:sigmaepsilonhat}&$O(Tn_x^2n_y)+O(Tn_xn_y^2)$\\
$\hat{\sigma}_{d}^2$&\eqref{eq:sigmadhat}&$O(n_x^3)+O(n_{\theta}^2)$\\

@@

The correlation analysis described in Section \ref{subsec:CorrAnal} were applied to generated data in order to approximate the unknown supports of both connectivity kernel and the disturbance covariance function. In the algorithm, the unknown variables $\sigma_{\epsilon}^2$ and $\xi$ in equations () and () were set to zero. By doing so, the exact shape of the kernel and disturbance covariance function cannot be inferred, this step might be considered as a prerequisite for the EM algorithm, however, an approximate of their support can be still extracted. In fact, the estimated support for the disturbance covariance function was very accurate. The estimates of the connectivity kernel and disturbance support were normalised and thresholded for values less than 0.01. The result of the correlation analysis for the connectivity kernel is shown in Figure.\ref{fig:KernelWidthEstimation}. The red circle shows the estimated support which encompasses the non-zero values (black area) after thresholding. The estimated support of the disturbance covariance function convolved twice with the observation kernel is shown in Figure.\ref{fig:DisturbanceWidthEstimation}. By fitting a Guassian basis function to the result, and under the assumption that the width of the sensor,$\sigma_m^2$ were known, the width of the disturbance covariance function, $\sigma_d^2$ was calculated analytically using the expression for Gaussian basis functions convolution.

The estimated supports were then applied in the EM algorithm described in Section \ref{sec:EstimationAlgorithm} to estimate the smoothed states $\hat{\mathbf x}_t^b$, the connectivity kernel weights $\boldsymbol\theta$, the synaptic dynamic $\xi$, and the hyperparameters $\sigma_{\epsilon}^2$ and $\sigma^2_d$. A  square grid of $n_{\theta}=25$ (5 by 5) basis functions with $2.25~mm$ spacing were used to cover the estimated support where $\sigma_{\psi}^2=3.5$. This is shown in Figure.\ref{fig:KernelWidthEstimation} by blue circles, each of which representing full width at half maximum of the connectivity kernel basis functions. The true and estimated connectivity kernel for 100 Monte Carlo simulations are shown in Figure~\ref{fig:KernelEstimates}. Cross-sections of the connectivity kernel estimates for each run are shown in Figure~\ref{fig:KernelEstimateCrossSection}. The reconstructed kernel is in good accordance with the actual kernel, where the actual kernel is inside the confidence interval. The large standard deviation is due to the high number of parameters need to be estimated. Figure~\ref{fig:Histogram} depicts the histogram of parameter estimates from 100 simulations where the true parameter value is shown by the solid line and the mean parameter estimate is shown by the dashed line. These histograms indicate small bias in parameter estimates with a very small standard deviation. Examples of neural field reconstruction are given in Figure.~\ref{fig:FieldEstimates}.  The figure shows snapshots at three time instants of the true neural field and the estimated field. In each subplot neural activity is plotted against spatial index in 1D to provide a comparision between true and the estimated underlying fields. Plot of lower bound of the log likelihood versus iterations of the EM is shown in Figure.~\ref{fig:LowerBound}. In general, for the non-linear dynamical systems the actual lower bound can not be claculated analytically, therefore the approximate of the lower bound obtained by URTSS is shown. The change in lower bound drops below 0.009 percent after 15 iterations, confirming
that the algorithm converged to steady parameter values.
\section{Discussion}
